Correlation among smart grid actors

### 2015-06-22 Article.tex
intro modif

Designing stable power systems is a classical engineering challenge since blackout events can have catastrophic consequences. A recent approach introduced by complex systems theorists consists in abstracting the power grid as a graph where nodes represent loads, generators, or transformers, while edges stand for electrical lines \cite{Pagani2012} \cite{Rosas-Casals2007} \cite{Pahwa}. As graphs, power grids display some characteristics like clustering coefficients, degree distributions, or density for instance. These metrics enable scientists to get a better understanding of the structure of such complex systems \cite{Pagani2012}. It appears that power grids are particular graphs compared to well-known large networks like the internet, the world wide web, or social networks. That is, they exhibit different characteritics, suggesting that they may behave differently \cite{Rosas-Casals2007}. A common hypothesis for such a difference is that power grids are completely designed by a very small number of entities (electricity operators mainly), whereas other pre-cited graphs results from the interactions of billions of independant entities. Therefore, properties like power law degree distributions, prerential attachment, or hub oriented attacks do not necessarily apply with power grids.

Actually, drawing conclusions based only on the study of the underlying graph structure may lead to unrealistic results \cite{Pahwa}. Power grids are ideed dynamical systems governed by physical laws. Concepts such as shortest path or centrality metrics have thus to be extended to take into account the particular nature of power grids. Besides, stability and resilience of such systems are much more complex than the simple connectivity of the underlying graph \cite{Pahwa}. It is well-known for instance that frequency synchronization in the grid is a necessary condition for stability. A common solution consists in modeling the grid as a network of coupled oscillators governed by a swing equation dynamic \cite{Filatrella2008}. It can then be shown that the topology of the network impacts directly the capacity of the oscillators to reach the synchronized state \cite{Dorfler2013}.

Understanding power grids seems like a reasonable starting point for studying so-called \textit{smart grids} that gained a lot of attention recently. As there exists many good articles \cite{Ramchurn} presenting smart grid ideas, concepts, or architectures, we will not dwell too much on this point. In this paper, we rather concentrate on a different approach to the system stability. We consider a set of agents that are equiped with distributed energy ressources (DER) as well as electrical loads, such that an agent can produce and consume energy at any time. His production can be, of course, used to meet his own demand, but in cases where he is over-producing, we consider that he has the possibility to sell his extra-production to the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper).

The constraints necessary for the electrical grid to remain in a stable state obiously forbid any anarchic system where prosumers inject power whenever and however they want. An envisaged solution for organizing such a system consists in a market environement, where participating entities announce a production capacity for an upcoming period of time. If the contract is bought, the concerned entity commits to injecting exactly at any time the contracted amount of power under financial penalities if it fails.

@@

We will see that variability in the aggregated productions can be quantified to a certain extend by the correlation among the agents forming the coalitions. Understanding the correlation reliationships among the agents can thus be illuminating in the sense that it gives an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different agents aggregations. We build a framework in which the system operator has the possibility of fixing some entrance conditions on the market both in terms of stability and sufficient production. Based on predictions, coalitions can decide wether or not to enter and what power quantities they are willing to provide. Of course, coalitions failing at fulfilling their obligations during the contracted period of time (because of bad forecasts, unexpected rare events, or lying) will be exposed to financial penalities.

Because agents are susceptible to fail for diverse reasons, the impact on the system's stability is critical. The propensy of a system to undertake these failures is often refered to as its resilience \cite{Pahwa}. Once formed, we also investigate in this paper the resilience of the coalitions when prosumers fail. Despite the fact that loosing agents is usually detrimental to the coalitions, we will see that, with our algorithm, coalitions tend to be less impacted by random failures.

@@

The intermittent nature of renewables such as wind or solar power introduces new challenges in control design. On the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the natural ressource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. A first approach to remedy this situation by acting on the consumption component is the use of dynamic ectricity prices sent to the end users. Together with demand side management techniques implemented on the smart meters, they consitute a first tool for shaping the consumption profiles \cite{Milligan2010}. In periods where the production is expected to be greater than the consumption, prices are scheduled low in order to encourage end users to delay their elastic loads to these periods. On the contrary, when the production is expected to be less than the consumption, high prices are scheduled to deter any delayable loads over these periods.

@@

This problem exhibits similarities with our objective of forming stable coalitions since our expected return can be formulated as the expected production and the risk / volatility as the variance of the production. However, it is assume in the portfolio theory \cite{Markowitz1959} that returns are normaly distributed (or at least that the joint distribution is elliptic). In this paper, we chose to use normal distributions only for explanation purposes, or for estimating some parameters. Therefore, all the results are computed based on observed distributions (sometines far from gaussians).

@@

Weather quantities like wind speed or solar irradiance appear thus as alternative data for generating the $ P_{i} $ series. Fortunately, these kind of data are easier to find, and since the developement of small personal weather stations, their geographical granularity keeps increasing.  A quantity like wind speed depends both on time and location (we neglect height since this information is rarely present in the data). In the following, time is discretized into slots and space into zones. A zone is simply a portion of the considered space for which we have weather data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

@@

\begin{figure}
\includegraphics[scale=.6]{mean_field.pdf}
\caption{Gaussian mean approximation. Subplot \textbf{a} shows how the parameter $ \alpha $ of the utility function should be chosen in function of the mean desired size of the coalitions (see equation \ref{eq:alpha_star}). Subplot \textbf{b} dislays the corresponding utility functions for different values of $ \alpha $. Blue curve with diamonds favors very small coalitions of 2 agents while the green one with triangles favors 5 agents coalitions. Finally, the red curve with squares has an optimum size of 15 agents. }
\label{fig:mean_approx}
\end{figure}

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not a study of coalitions stability against player defection, wich game theory provides a lot of tools for. This is indeed a problem on its own. The redistribution of a coalition's utility in terms of individual payoffs will therefore not be considered directly in this paper. $ \mathcal{U} $ can be rather interpreted as a measure of how good a given coalition is according to our criteria since it favors coalitions with good production to risk ratios.

@@

Before running the algorithms, we need to calibrate the utility function by choosing the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitely considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utitility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to estimate $ \alpha $ based on the mean quantities and gaussian approximations seems to give acceptable results for the utility function behavior.

@@

In this paper we studied how aggregations of prosumers could be authorized to sell their surplus of production to the grid operator. By relying on the past values of the agents, we constrained the market entry to both sufficiently productive and stable coalitions. The power that a coalition is able to propose on the market is therefore related to both quantities. As the correlations between the prosumers that form these coalitions impact directly their volatilities, we seek uncorrelated aggregations of agents. We used a graph representation of the correlation relationships between the agents as a reduced landscape for the coalition formation. A greedy algorithm that starts with cliques of the "de-correlation" graph of the agents and makes local improvements offers a good compromise between speed and quality of the results. We compare these results with a random search, and an opposite strategy that clusters correlated agents together. We showed that the coalitions resulting from our algorithm are able to provide more power to the grid with a lower volatility. Because they tend to have globally better production over volatility ratios, these coalitions will tend to use les storage and spill less energy than more unstable ones. We plan to study these benefits for the control of the aggregations in future works.

>>>>

%Designing stable power systems is a classical engineering challenge since blackout events can have catastrophic consequences. A recent approach introduced by complex systems theorists consists in abstracting the power grid as a graph where nodes represent loads, generators, or transformers, while edges stand for electrical lines \cite{Pagani2012} \cite{Rosas-Casals2007} \cite{Pahwa}. As graphs, power grids display some characteristics like clustering coefficients, degree distributions, or density for instance. These metrics enable scientists to get a better understanding of the structure of such complex systems \cite{Pagani2012}. It appears that power grids are particular graphs compared to well-known large networks like the internet, the world wide web, or social networks. That is, they exhibit different characteritics, suggesting that they may behave differently \cite{Rosas-Casals2007}. A common hypothesis for such a difference is that power grids are completely designed by a very small number of entities (electricity operators mainly), whereas other pre-cited graphs results from the interactions of billions of independant entities. Therefore, properties like power law degree distributions, prerential attachment, or hub oriented attacks do not necessarily apply with power grids.

%Actually, drawing conclusions based only on the study of the underlying graph structure may lead to unrealistic results \cite{Pahwa}. Power grids are ideed dynamical systems governed by physical laws. Concepts such as shortest path or centrality metrics have thus to be extended to take into account the particular nature of power grids. Besides, stability and resilience of such systems are much more complex than the simple connectivity of the underlying graph \cite{Pahwa}. It is well-known for instance that frequency synchronization in the grid is a necessary condition for stability. A common solution consists in modeling the grid as a network of coupled oscillators governed by a swing equation dynamic \cite{Filatrella2008}. It can then be shown that the topology of the network impacts directly the capacity of the oscillators to reach the synchronized state \cite{Dorfler2013}.

Designing stable power systems is a classical engineering challenge since blackout events can have catastrophic consequences. One natural condition for stability revolves around sustaining a power generation that meets the demand at any time. If one is larger than the other, the system deviates from its synchronous state. If nothing is done to return to the synchronized equilibrium, this can lead to catastrophic cascades of failures. With traditional power plants based on fossil energies, the production can be scheduled in order to sustain the predicted consumption. Relatively small deviations from this schedule can then be supported almost on real time by using fast power plants, interconnections with border countries, as well as regulator entities. Even if most individual consumers may not realize it, the electicity prices are not constant rates and evolve with the production/consumption conditions on the grid. Operations are decided on a market where contracts are created based on information from the grid. Such a market environment appears necessary in order to improve the reliability of the whole system, and is already being used with large renewable plants in the production portfolio.

The smart grid vision however goes further and revolutionalizes this top-down centralized architecture by assuming that bidirectional power flows are allowed. This would change completely the nature of the traditional grid since production could be located down to the very end of the distribution network. Nodes that were simply pure loads yesterday could behave tomorow as generators or loads depending on time. On the other hand, the penetration of renewables in the production is continuously growing, and is expected to become majority in a near future. These plants completely rely on the presence and the intensity of their respective ressources (wind, sun, tiles...), which predictions can be difficult and uncertain. Balancing production and demand in such a scenario appears much more challenging and it seems inevitable that the grid should modernize its infrastructures to sustain this transition. The key to drive safely such a system is assumed to be information, and more precisely, the capacity to measure, communicate, and analyze data on real time.

%Understanding power grids seems like a reasonable starting point for studying so-called \textit{smart grids} that gained a lot of attention recently. As there exists many good articles \cite{Ramchurn} presenting smart grid ideas, concepts, or architectures, we will not dwell too much on this point.

In this paper, we focus on these nodes in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that can be seen as aggregations of renewable distributed energy ressources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases where it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper).

The constraints necessary for the electrical grid to remain in a stable state obiously forbid any anarchic system where prosumers inject power whenever and however they want. In this paper, the envisaged solution for organizing such a multi-producers system consists in a market environement, where participating entities announce an estimated production capacity for an upcoming period of time. If a contract is bought, the concerned entity commits to injecting exactly, at any time of the contracted period, the announced amount of power, under financial penalities if it fails.

@@

We will see that variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation reliationships among the agents can thus be illuminating in the sense that it gives an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different agents aggregations. We build a framework in which the system operator has the possibility of specifying both a maximum amount of risk and a minimum production acceptable to enter the market. Based on predictions, coalitions can decide wether or not to enter and what power quantities they are willing to provide. Of course, coalitions failing at fulfilling their obligations during the contracted period of time (because of bad forecasts, unexpected rare events, or intentional false information) will be exposed to financial penalities.

Because agents are susceptible to fail for diverse reasons, the impact on the system's stability is critical. The propensy of a system to undertake these failures is often refered to as its resilience \cite{Pahwa}. Once formed, we also investigate in this paper the resilience of the coalitions when prosumers fail. Despite the fact that losing agents is usually detrimental to the coalitions, we will see that, with our algorithm, coalitions tend to be less impacted by random failures.

@@

The intermittent nature of renewables such as wind or solar power introduces new challenges in control design. On the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the natural ressource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. A first approach to remedy this situation by acting on the consumption component is the use of dynamic ectricity prices sent to the end users. Together with demand side management techniques implemented on the smart meters, they consitute a first tool for shaping the consumption profiles \cite{Milligan2010}. In periods when the production is expected to be greater than the consumption, prices are scheduled low in order to encourage end users to delay their elastic loads to these periods. On the contrary, when the production is expected to be less than the consumption, high prices are scheduled to deter any delayable loads over these periods.

@@

This problem exhibits similarities with our objective of forming stable coalitions since our expected return can be formulated as the expected production and the risk / volatility as the variance of the production. However, it is assumed in the portfolio theory \cite{Markowitz1959} that returns are normaly distributed. In this paper, we chose to use normal distributions only for explanation purposes, or for estimating some parameters. Therefore, all the results are computed based on observed distributions (sometines far from gaussians).

@@

Weather quantities like wind speed or solar irradiance appear thus as alternative data for generating the $ P_{i} $ series. Fortunately, these kinds of data are easier to find, and since the developement of small personal weather stations, their geographical granularity keeps increasing.  A quantity like wind speed depends both on time and location (we neglect height since this information is rarely present in the data). In the following, time is discretized into slots and space into zones. A zone is simply a portion of the considered space for which we have weather data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

@@

\begin{figure}
\includegraphics[scale=.6]{mean_field.pdf}
\caption{Gaussian mean approximation. Subplot \textbf{a} shows how the parameter $ \alpha $ of the utility function should be chosen in function of the mean desired size of the coalitions (see equation \ref{eq:alpha_star}). Subplot \textbf{b} dislays the corresponding utility functions for different values of $ \alpha $. Blue curve with diamonds favors very small coalitions of 2 agents while the green one with triangles favors 5 agents coalitions. Finally, the red curve with squares has an optimal size of 15 agents. }
\label{fig:mean_approx}
\end{figure}

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not a study of coalitions stability against player defection, which game theory provides a lot of tools for. This is indeed a problem on its own. The redistribution of a coalition's utility in terms of individual payoffs will therefore not be considered directly in this paper. $ \mathcal{U} $ can be rather interpreted as a measure of how good a given coalition is according to our criteria since it favors coalitions with good production to risk ratios.

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitely considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utitility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to estimate $ \alpha $ based on the mean quantities and gaussian approximations seems to give acceptable results for the utility function behavior.

@@

In this paper we studied how aggregations of prosumers could be authorized to sell their surplus of production to the grid operator. By relying on the past values of the agents, we constrained the market entry to both sufficiently productive and stable coalitions. The power that a coalition is able to propose on the market is therefore related to production and stability. As the correlations between the prosumers that form these coalitions impact directly their volatilities, we seek uncorrelated aggregations of agents. We used a graph representation of the correlation relationships between the agents as a reduced landscape for the coalition formation. A greedy algorithm that starts with cliques of the "de-correlation" graph of the agents and makes local improvements offers a good compromise between speed and quality of the results. We compare these results with a random search, and an opposite strategy that clusters correlated agents together. We showed that the coalitions resulting from our algorithm are able to provide more power to the grid with a lower volatility. Because they tend to have globally better production over volatility ratios, these coalitions will tend to use less storage and spill less energy than more unstable ones. We plan to study these benefits for the control of the aggregations in future works.
